---
title: "Visualization Lecture"
author: "Kaleb J. Frierson"
date: "2024-09-26"
output: 
  github_document: 
    toc: TRUE
---

# Lecture Notes

Exploratory data analysis can mean a lot of things. You're getting to know your data! Everything you have done after importing and tidying data up until you do official testing/analyses.

Looking at data is critical. You will make dozens of graphics for each dataset. You will be the only person who views a lot of them. A small subset are for others, but visualizing your data is crucial throughout your work.

ggplot makes good graphics with relative ease (compared to base R graphics). gg = grammar of graphics. Conceptual framework for how you want to construct plots. Requires some learning about what that grammar is. But once you know it, the infrastructure is the same.

Learning ggplot: lots of online materials. Google is helpful! David Robinson: "Just discovered that I've visited this ggplot2 question 29 times in the last two years."

Basic graph components: data, aesthetic mappings, geoms

Advanced graph components: facets, scales, statistics

A graph is built by combining these components. Components are consistent across graph types!

# Coding

## libraries and setup

```{r setup, include = FALSE}
library(tidyverse)
library(ggridges)
```

```{r, warning=FALSE}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

## intro to ggplot

```{r}
weather_df

```

Making first plot, note that unless you assign geometry you won't get data on the plot.

```{r}
ggp_weather_scatterplot = 
  ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

ggp_weather_scatterplot

```

Alpha can change shading of points. Color by name adds a key. Geom smooth fits a smooth line and defaults with standard error which you can remove with SE = FALSE.

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y=tmax, color = name)) +
  geom_point(alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

Where you define aesthetics matters:

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y=tmax)) +
  geom_point(aes(color = name), alpha = .3, size = .8) + 
  geom_smooth(se = FALSE)
```

use faceting real quick:

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y=tmax, color = name)) + 
  geom_point(alpha=0.3) + 
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name) #this is written such that columns are first, you can also do it the other way arounf (name ~ .)

```

## Scatter plots

Let's make a somewhat more interesting scatterplot:

```{r}
weather_df |> 
  ggplot(aes(x= date, y= tmax, color = name, size = prcp)) + 
  geom_point(alpha = 0.3) + 
  geom_smooth(se = FALSE) + 
  facet_grid(.~name)
```

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  mutate(
    tmax_fahr = tmax * (9 / 5) + 32,
    tmin_fahr = tmin * (9 / 5) + 32) |> 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = .5) + 
  geom_smooth(method = "lm", se = FALSE, color = "green")
```
## Small things

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y=tmax))+
  geom_point(aes(color = name), alpha = .3, size= .8) + 
  geom_smooth(se = FALSE)

```

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y=tmax))+
  geom_hex()
```
```{r}
weather_df |> 
  ggplot(aes(x= tmin, y=tmax)) +
  geom_point(color = "royalblue")
```

Colors bring up a great question. When to quote and when not to quote? 

Generally speaking, if something is a variable name, it doesn't need to be in quotes. If it is NOT a variable name it often needs quotes. Unless of course it is a function. 


## histograms

color defines the outside of the histogram bar, fill will fill the bar. It is still difficult to compare these things. You can say in histogram position = dodge and it will put bars next to eachother instead of on top of eachother. 

```{r}
weather_df |> 
  ggplot(aes(x=tmin, color = name)) + 
  geom_histogram(position="dodge")
```
Still difficult to see, how to fix? Useing facet? 

```{r}
weather_df |> 
  ggplot(aes(x=tmin, fill=name)) + 
  geom_histogram()+
  facet_grid(.~name)
```

This is kind of clunky. Maybe not the best way to compare the data. Great for seeing distributions. 

Density plot? Get a more "smooth" histogram. Helps make comparisons and see the distributions. 

```{r}
weather_df |> 
  ggplot(aes(x=tmin, fill = name))+
  geom_density(alpha = 0.3)
```

## box plots

Color not appropriate here for identifying but might help for consistency if you decide to report these graphs or you just like pretty colors: 
```{r}
weather_df |> 
  ggplot(aes(x=name, y=tmin, fill = name)) + 
  geom_boxplot()

```
## violin plots

```{r}
weather_df |> 
  ggplot(aes(x=name, y=tmin, fill = name)) + 
  geom_violin()
```
## ridge plots

didn't exist and then in 2017 they went crazy

```{r}
weather_df |> 
  ggplot(aes(x=tmin, y= name)) + 
  geom_density_ridges()
```
## Learning assessment

```{r}
ggplot(weather_df, aes(x = prcp)) + 
  geom_density(aes(fill = name), alpha = .5) 
```

```{r}
ggplot(weather_df, aes(x = prcp, y = name)) + 
  geom_density_ridges(scale = .85)
```

```{r}
ggplot(weather_df, aes(y = prcp, x = name)) + 
  geom_boxplot() 
```

```{r}
weather_df |> 
  ggplot(aes(x=name, y=prcp, fill = name)) + 
  geom_violin()
```

Lots of days there is no rain, lets filter to make graphics better:

```{r}
weather_df |> 
  filter(prcp > 0) |> 
  ggplot(aes(x = prcp, y = name)) + 
  geom_density_ridges(scale = .85)

```

```{r}
weather_df |> 
  filter(prcp > 10) |> 
  ggplot(aes(x = prcp, y = name)) + 
  geom_density_ridges(scale = .85)

```

What if you have one outlier that is really stretching you axis? Take it out if you want! 

## saving and embedding plots

```{r, warning= FALSE}
ggp_weather= 
  weather_df |> 
  ggplot(aes(x= date, y = tmax, color = name)) + 
  geom_point() 

ggsave("plots/ggp_weather.pdf", ggp_weather, width = 8, height = 6)

```

```{r embed, fig.width = 6, fig.asp = .6}
  weather_df |> 
  ggplot(aes(x= date, y = tmax, color = name)) + 
  geom_point() 
```

